2.1 APL Data Quality Check with additonal cleaning

### 2025-07-07 2_scripts/2.1 APLv1 Data Cleaning.qmd
Removed unnecessary scripts

# Second version of APL: provided by Carmen in Jan 2024

>>>>

# Second version of APL: provided by Carmen in Jan 2024
library(readxl)
RFD4682_v2 <- read_excel("../1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
    sheet = "RFD4682_Full_Data", col_types = c("text", 
        "text", "text", "text", "text", 
        "date", "text", "text", "date", "date", 
        "text", "text", "numeric", "text", 
        "text", "text", "text", "text", "numeric", 
        "numeric", "text", "text", 
        "numeric", "numeric", "numeric"))
View(RFD4682_v2)
```

### 2025-07-07 2_scripts/2.1.1 APLv2 Data Cleaning.qmd
Removed unnecessary scripts

Estimation on number of individuals, using cleaned version of IDe
```{r}
# Total num of individual in the data:
table(is.na(unique(RFD4682_e$clean_IDe)), useNA = 'ifany')    # N=188,722, plus 1 NA

@@

Some additional cleaning/editing, making anti-S quantitative
```{r}
# Missing values
summary(is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F) 
# N=36, beware of "not available" values, better recode those as NAs

@@

##### Yuan found this: anti-N is only available for the Omicron stage
**Notes:** This is an error in two ver of APL. The anti-N assay results are available and misplaced in the anti-S-RBD column.
```{r}
# The APL Data:
RFD4682_e$COLLECTION_DATE %>%
  summary()

@@

#### ADD order ID and other ids -- doesn't work
```{r}
RFD4682_v2$order_ID<-RFD4682_e$order_ID

@@

#### 2.1 De-duplicate
```{r}
# Does duplicated records exist?
table(duplicated(RFD4682_v2))  # no-all records are unique

@@

RFD4682_v2<-rbind.data.frame(RFD4682_v2, RFD4682_v2dup2[c(2,4,6,8,9:16),])
rm(RFD4682_v2dup, RFD4682_v2dup2)
```
##### De-duplicate SERO_ID for APL v1
```{r}
# use the saved duplicated SERO_ID variables to de_duplicate:
RFD4682_edup<-RFD4682_e %>%
 .[duplicated(RFD4682_e$SERO_ID)==T,]

@@

#### 2.2 Add ULI_IDe from v1
```{r}
library(duawranglr)    # use this to encrypt the ids

@@

```{r}
# Use SERO_ID and 'RBD-IgGII_RESULT' to identify each records
RFD4682_v2<-merge(RFD4682_v2, select(RFD4682_e, SERO_ID, order_ID, PERSON_IDe, ULI_IDe,
                         clean_IDe, QUINTMAT, QUINTSOC), 
      by ="SERO_ID", all = T
      )
# ULI - ULI_IDe crosswalk
ULI_crosswalk<-select(RFD4682_v2, ULI_ID, ULI_IDe)
ULI_crosswalk<-distinct(ULI_crosswalk)

table(is.na(ULI_crosswalk$ULI_IDe))
ULI_crosswalk %>%
  .[is.na(.$ULI_IDe),]

# APL ver2 seems have solved missing value issue in the ULI ID, needs to add those
table(is.na(RFD4682_v2$ULI_ID), is.na(RFD4682_v2$ULI_IDe))

table(is.na(RFD4682_v2$PERSON_ID), is.na(RFD4682_v2$PERSON_IDe))

# encrypt new ULI_ID
set.seed(240611)
ULI_crosswalk_e <- deid_dua(ULI_crosswalk, 
                      id_col = 'ULI_ID',
                      new_id_name = 'ULI_IDe2',
                      write_crosswalk = F, id_length = 15)

ULI_crosswalk$ULI_IDe2<-ULI_crosswalk_e$ULI_IDe2

# clean the ULI_ID crosswalk file
table(is.na(ULI_crosswalk$ULI_ID))
ULI_crosswalk<-ULI_crosswalk[!is.na(ULI_crosswalk$ULI_ID),]
ULI_crosswalk<-ULI_crosswalk[-64591,]
ULI_crosswalk$ULI_IDe[is.na(ULI_crosswalk$ULI_IDe)]<-ULI_crosswalk$ULI_IDe2[is.na(ULI_crosswalk$ULI_IDe)]

# apply the cross-walk to the v2 data
df.temp1<-RFD4682_v2[!is.na(RFD4682_v2$ULI_ID) & is.na(RFD4682_v2$clean_IDe),]
df.temp2<-RFD4682_v2[!(!is.na(RFD4682_v2$ULI_ID) & is.na(RFD4682_v2$clean_IDe)),]

df.temp1<-merge(select(df.temp1, -c("ULI_IDe")), 
             ULI_crosswalk[,1:2],
      by="ULI_ID", all.x = T, all.y = F)
df.temp1$clean_IDe<-df.temp1$ULI_IDe

RFD4682_v2<-bind_rows(df.temp1, df.temp2)
rm(df.temp1, df.temp2, ULI_crosswalk_e)
```

```{r}
# Still, there's some missing value in clean_IDe
RFD4682_v2 %>%
  .[is.na(.$clean_IDe),]
```
#### 2.3 Clean anti-N and anti-S

##### 2.3.1 Anti-N cleaning (remove index)
```{r}
# Create an indicator for misplaced anti-N
RFD4682_v2$abb_s_index<-grepl("index", RFD4682_v2$`RBD-IgGII_RESULT`, ignore.case = T)
RFD4682_v2$abb_s_AU<-grepl("ml", RFD4682_v2$`RBD-IgGII_RESULT`, ignore.case = T)
table(RFD4682_v2$abb_s_index, RFD4682_v2$abb_s_AU, useNA = 'ifany')
# 36 samples missing abbott_s info

@@

##### 2.3.2 Anti-S cleaning (remove AU/ml)
```{r}
RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[a-z]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[A-Z]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[/<>]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
RFD4682_v2$`RBD-IgGII_RESULT`<-as.numeric(RFD4682_v2$`RBD-IgGII_RESULT`)
summary(RFD4682_v2$`RBD-IgGII_RESULT`)

@@

```
#### 2.4 Remove unencryoted individual inditifier
```{r}
RFD4682_ev2<-select(RFD4682_v2, -c(ULI_ID, PERSON_ID, misplace))
```

#### 2.5 Save the encrypted ver of APL data
```{r}
save(RFD4682_v2, file = "../1_data/private/RFD4682_v2.RData")
save(RFD4682_ev2, file = "../1_data/private/RFD4682_ev2.RData")

@@

## 3. Clean other two APL sub datasets (Vax, NAAT)

## Unfinished ##
```{r}
# Add encrypted IDe to vax and NAAT data
RFD4682_v2_vaxe<-merge(RFD4682_v2_vax, select(ULI_crosswalk, ULI_ID, ULI_IDe),
      by.x="ULI", by.y="ULI_ID", all.x = T, all.y = F)
# any missing?
table(is.na(RFD4682_v2_vaxe$ULI_IDe))

# also encrypt the ULI_ID in the RFD4682_vax

list_temp<-RFD4682_v2_vaxe %>%
  .[is.na(.$ULI_IDe),]
```

>>>>

Estimation on number of individuals, using cleaned version of IDe

```{r}
# Total num of individual in the data:
table(is.na(unique(RFD4682_e$clean_IDe)), useNA = 'ifany')    # N=188,722, plus 1 NA

@@

Some additional cleaning/editing, making anti-S quantitative

```{r}
# Missing values
summary(is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F) 
# N=36, beware of "not available" values, better recode those as NAs

@@

##### Yuan found this: anti-N is only available for the Omicron stage

**Notes:** This is an error in two ver of APL. The anti-N assay results are available and misplaced in the anti-S-RBD column.

```{r}
# The APL Data:
RFD4682_e$COLLECTION_DATE %>%
  summary()

@@

#### ADD order ID and other ids -- doesn't work

```{r}
RFD4682_v2$order_ID<-RFD4682_e$order_ID

@@

#### 2.1 De-duplicate

```{r}
# Does duplicated records exist?
table(duplicated(RFD4682_v2))  # no-all records are unique

@@

RFD4682_v2<-rbind.data.frame(RFD4682_v2, RFD4682_v2dup2[c(2,4,6,8,9:16),])
rm(RFD4682_v2dup, RFD4682_v2dup2)
```

##### De-duplicate SERO_ID for APL v1

```{r}
# use the saved duplicated SERO_ID variables to de_duplicate:
RFD4682_edup<-RFD4682_e %>%
 .[duplicated(RFD4682_e$SERO_ID)==T,]

@@

#### 2.2 Add ULI_IDe from v1

```{r}
library(duawranglr)    # use this to encrypt the ids

@@

```{r}
# Use SERO_ID and 'RBD-IgGII_RESULT' to identify each records
RFD4682_v2<-merge(RFD4682_v2, select(RFD4682_e, SERO_ID, order_ID, PERSON_IDe, ULI_IDe,
                         clean_IDe, QUINTMAT, QUINTSOC), 
      by ="SERO_ID", all = T
      )
```

```{r}
# Still, there's some missing value in clean_IDe
RFD4682_v2 %>%
  .[is.na(.$clean_IDe),]
```

#### 2.3 Clean anti-N and anti-S

##### 2.3.1 Anti-N cleaning (remove index)

```{r}
# Create an indicator for misplaced anti-N
RFD4682_v2$abb_s_index<-grepl("index", RFD4682_v2$`RBD-IgGII_RESULT`, ignore.case = T)
RFD4682_v2$abb_s_AU<-grepl("ml", RFD4682_v2$`RBD-IgGII_RESULT`, ignore.case = T)
table(RFD4682_v2$abb_s_index, RFD4682_v2$abb_s_AU, useNA = 'ifany')
# 36 samples missing abbott_s info

@@

##### 2.3.2 Anti-S cleaning (remove AU/ml)

```{r}
RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[a-z]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[A-Z]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[/<>]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
RFD4682_v2$`RBD-IgGII_RESULT`<-as.numeric(RFD4682_v2$`RBD-IgGII_RESULT`)
summary(RFD4682_v2$`RBD-IgGII_RESULT`)

@@

#### 2.4 Remove unencryoted individual inditifier

```{r}
RFD4682_ev2<-select(RFD4682_v2, -c(ULI_ID, PERSON_ID, misplace))
```

#### 2.5 Save the encrypted ver of APL data

```{r}
save(RFD4682_v2, file = "../1_data/private/RFD4682_v2.RData")
save(RFD4682_ev2, file = "../1_data/private/RFD4682_ev2.RData")

@@
